---
title: "LaTeX memoir Customization"
output: 
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
vignette: >
  %\VignetteIndexEntry{LaTeX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
set.seed(97310)
```

The *Memoir* template is based of the LaTeX **memoir** package.
It is optimized for long documents and is very versatile, with very different possible layouts.

The template can be highly customized by simple changes in the header of `index.Rmd`.
This short article presents the main features of the customization.
The details to implement them, namely the options in the header of `index.Rmd` are in the template text.

Two main choices must be made:

- The chapter and page styles.
Many are proposed by **memoir**.
- The way margins are used. 
Large margins host sidenotes rather than footnotes, bibliographic references and figure captions. 
Small margins remain empty.

The default options produce a small-margin document, with an elaborate chapter style.
It is presented in the [gallery](https://ericmarcon.github.io/memoiR/gallery/memoir/bookdown_pdf_book/MyBook.pdf).
It can be changed quickly for a [wide-margin layout](https://ericmarcon.github.io/travailleR/travailleR.pdf), whose parameters are given in comments.


# Chapter and page styles

Chapter styles define the way top-level titles are presented.
A [gallery](https://mirrors.ctan.org/info/latex-samples/MemoirChapStyles/MemoirChapStyles.pdf) is available.

The default chapter style of the template is `daleif1`.
It is not actually supported natively by **memoir** but its code is added in `preamble.tex`.
It proposes graphic chapter titles.
It works well with the  `Ruled` page style, with a page header that recalls the chapter and section titles.
All page styles are described in the section 7.2 of the **memoir** [manual](https://mirrors.ctan.org/tex-archive/macros/latex/contrib/memoir/memman.pdf).

An alternative is `companion`, both chapter and page style employed in the large margin layout template.


# Margins

A4 or Letter paper stock with 1-inch margin on both sides imply long lines, actually too long to read comfortably.
Avoid lines other 80 characters by increasing margin width.
Two options are used:

```
smallmargin: 1.5in         # outer margin (small).
spinemargin: 1.5in         # spine margin. Reduce to 1.2 with large margins.
```

Large margin allow both a narrower text column for readability and putting stuff into them.
Use `largemargins: true` to use them.
The outer margin is the set to `largemargin: 3in` and the spine margin may be reduced a bit.


# Bibliographic styles

Many bibliographic styles come with **biblatex**, the LATeX package in charge of reference management.
A [gallery](https://fr.overleaf.com/learn/latex/Biblatex_bibliography_styles) is available.

The style is selected in the options of biblatex:

```
biblatexoptions:
  - style=verbose-inote
  - pageref=true
```

The default style of the template is a classical author-year one, with the ibidem option, i.e.the authors are not repeated if they are the same as the previous reference.
When large margins are available, the `verbose-inote` style is appropriate for the reader to have the full references close to the text, including a DOI link.

More styles can be used, including specific styles requiring an extra LaTeX package that is usually installed automatically by LaTeX distributions.
The `pageref` option must be commented out if the chosen style does not support it or an error may occur during the LaTeX compilation of the document.


# Fonts

The default font of the memoir is Computer Modern, the LaTeX standard.

Other fonts may be used.
They must be installed in a LaTeX package, such as [**tex-gyre**](https://www.ctan.org/pkg/tex-gyre) and its companion **tex-gyre-math**.
Both provide updated versions of the classical:

- Times New Roman: `Tex Gyre Termes`
- Century Schoolbook: `Tex Gyre Schola`
- Palatino: `Tex Gyre Pagella`
- Helvetica: `Tex Gyre Heros`
- Courier: `Tex Gyre Cursor`
- Bookman Old Style: `Tex Gyre Bonum`
- Avant Garde Gothic: `Tex Gyre Adventor`
- Zapf Chancery: `Tex Gyre Chorus`

The LaTeX packages must be installed in the LaTeX distribution.
If **tinytex** installed it, it can do that too:
```{r, eval=FALSE}
tinytex::tlmgr_install(c("tex-gyre", "tex-gyre-math"))
```

This command must be added to the GitHub Action script created by `build_ghworkflow()`, just after `tinytex::install_tinytex()`.

The fonts are declared in the header of `index.Rmd`.
The math font name is the same as the main font completed with `Math`.

```
mainfont: TeX Gyre Schola
mathfont: TeX Gyre Schola Math
```

The main font is a serif font, here Century Schoolbook.
The monotype font is used in code blocks, and in code included in the text.
It may be `monofont: TeX Gyre Cursor`, but the default font usually is usually better.

Note that long tables (`longtable = TRUE` argument in the `kable()` function to render tables) may not be cut correctly accross pages when specific fonts are declared.


# Conclusion

Large margins are optimal with the `companion` style and the `verbose-inote` bibliographic style; small margins should be used with the default options of the template.
Much of memoir customization is offered by the template by choosing other styles, and possibly other fonts.
