---
title: "LaTeX memoir Customization"
output: 
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
vignette: >
  %\VignetteIndexEntry{LaTeX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
set.seed(97310)
```

The *Memoir* template is based of the LaTeX **memoir** package.
It is optimized for long documents and is very versatile, with very different possible layouts.

The template can be highly customized by simple changes in the header of `index.Rmd`.
This short article presents the main features of the customization.
The details to implement them, namely the options in the header of `index.Rmd` are in the template text.

Two main choices must be made:

- The chapter and page styles.
Many are proposed by **memoir**.
- The way margins are used. 
Large margins host sidenotes rather than footnotes, bibliographic references and figure captions. 
Small margins remain empty.

The default options produce a small-margin document, with an elaborate chapter style.
It is presented in the [gallery](https://ericmarcon.github.io/memoiR/gallery/memoir/bookdown_pdf_book/MyBook.pdf).
It can be changed quickly for a [wide-margin layout](https://ericmarcon.github.io/travailleR/travailleR.pdf), whose parameters are given in comments.


# Chapter and page styles

Chapter styles define the way top-level titles are presented.
A [gallery](https://mirrors.ctan.org/info/latex-samples/MemoirChapStyles/MemoirChapStyles.pdf) is available.

The default chapter style of the template is `daleif1`.
It is not actually supported natively by **memoir** but its code is added in `preamble.tex`.
It proposes graphic chapter titles.
It works well with the  `Ruled` page style, with a page header that recalls the chapter and section titles.
All page styles are described in the section 7.2 of the **memoir** [manual](https://mirrors.ctan.org/tex-archive/macros/latex/contrib/memoir/memman.pdf).

An alternative is `companion`, both chapter and page style employed in the large margin layout template.


# Margins

A4 or Letter paper stock with 1-inch margin on both sides imply long lines, actually too long to read comfortably.
Avoid lines other 80 characters by increasing margin width.
Two options are used:

```
smallmargin: 1.5in         # outer margin (small).
spinemargin: 1.5in         # spine margin. Reduce to 1.2 with large margins.
```

Large margins allow both a narrower text column for readability and putting stuff into them.
Use `largemargins: true` to use them.
The outer margin is then set to `largemargin: 3in` and the spine margin may be reduced a bit.


# Bibliographic styles

Many bibliographic styles come with **biblatex**, the LaTeX package in charge of reference management.
A [gallery](https://fr.overleaf.com/learn/latex/Biblatex_bibliography_styles) is available.

The style is selected in the options of biblatex:

```
biblatexoptions:
  - style=verbose-inote
  - pageref=true
```

The default style of the template is a classical author-year style, with the ibidem option, i.e. the authors are not repeated if they are the same as in the previous reference.
When large margins are available, the `verbose-inote` style is appropriate for the reader to have the full references close to the text, including a DOI link.

More styles can be used, including specific styles requiring an extra LaTeX package that is usually installed automatically by LaTeX distributions.
The `pageref` option must be commented out if the chosen style does not support it or an error may occur during the LaTeX compilation of the document.

If verbose citations in large margins are wanted but another bibliographic style is desired, `style` can be replaced by two separate instructions for citation and bibliographic styles, such as:

```
biblatexoptions:
  - citestyle=verbose-inote
  - pageref=true
  - bibstyle=apa
```

Note that the apa style requires the **biblatex-apa** package installed in the LaTeX distribution.


# Fonts

The default font of the memoir is Computer Modern (actually, an update of it), the LaTeX standard.

Other fonts may be used: the [LaTeX font catalogue](https://tug.org/FontCatalogue/alphfonts.html) is their gallery.
They must be installed in a LaTeX package, such as [**tex-gyre**](https://www.ctan.org/pkg/tex-gyre) and its companion **tex-gyre-math**.
Both provide updated versions of the classical:

- Times New Roman: [`Tex Gyre Termes`](https://tug.org/FontCatalogue/texgyretermes/)
- Century Schoolbook: [`Tex Gyre Schola`](https://tug.org/FontCatalogue/texgyreschola/)
- Palatino: [`Tex Gyre Pagella`](https://tug.org/FontCatalogue/texgyrepagella/)
- Helvetica: [`Tex Gyre Heros`](https://tug.org/FontCatalogue/texgyreheros/)
- Courier: [`Tex Gyre Cursor`](https://tug.org/FontCatalogue/texgyrecursor/)
- Bookman Old Style: [`Tex Gyre Bonum`](https://tug.org/FontCatalogue/texgyrebonum/)
- Avant Garde Gothic: [`Tex Gyre Adventor`](https://tug.org/FontCatalogue/texgyreadventor/)
- Zapf Chancery: [`Tex Gyre Chorus`](https://tug.org/FontCatalogue/texgyrechorus/)

The LaTeX packages must be installed in the LaTeX distribution.
If **tinytex** installed it, it can do that too:
```{r, eval=FALSE}
tinytex::tlmgr_install(c("tex-gyre", "tex-gyre-math"))
```

This command must be added to the GitHub Action script created by `build_ghworkflow()`, just after `tinytex::install_tinytex()`.
`build_ghworkflow()` does it automatically if it recognizes the font, see below.

The fonts are declared in the header of `index.Rmd`.
XeLaTeX is not able to find fonts by their name if they are not installed at the system level, namely when installed by packages.
They must be declared by their file names, which are in general their names without spaces, in lower case.
The main font options allow building the actual file names and are not likely to be changed.

```
mainfont: texgyretermes
mainfontoptions:
  - Extension=.otf
  - UprightFont=*-regular
  - BoldFont=*-bold
  - BoldItalicFont=*-bolditalic
mathfont: texgyretermes-math.otf
```
The math font name is fully specified, with no options.

A monotype font may be declared to be used in code blocks, and in code included in the text.
It may be `monofont: texgyrecursor`, completed by an identical `monofontoptions` list, but the default font is usually better.

Note that long tables (`longtable = TRUE` argument in the `kable()` function to render tables) may not be cut correctly across pages when specific fonts are declared.

The font size is 12pt by default.
Change it if necessary, depending on the font type, the margin width and personal preferences.


# GitHub Actions workflow

The memoir can be produced by GitHub Actions and made visible in GitHub Pages by the `memoir.yml` script (in `.github\workflows`) created by `build_ghworklow()`.

The script installs tinytex, a low-weight TeXLive distribution, with the **tinytex** R package.
Neither additional fonts nor language support beyond English are installed by default.
`build_ghworklow()` adds the necessary commands in the script (basically, install LaTeX packages) by reading the fonts and languages options in the header of the document and recognizing the LaTeX packages they need.

Support is necessarily limited.
Supported languages and fonts are listed here.
Other ones can be used in a memoir, but the `memoir.yml` script will have to be edited manually to allow GitHub Actions to work properly.

## Supported languages

The hyphenation pattern of a language must be installed in the LaTeX distribution.
If it is not, text will just not be hyphenated.
Supported languages are:

- `de`: German,
- `fr`: French,
- `it`: Italian,
- `pt`: Portuguese,
- `sp`: Spanish.

English is supported natively.

## Supported fonts

Font file names are used to find the corresponding package and install it.
For instance, `mainfont: texgyretermes`, as all fonts whose name starts with `texgyre` needs the **tex-gyre** LaTeX package.

Supported font families are listed here with their name (the file name pattern) and their LaTeX package:

- DejaVu (`dejavu*`): **dejavu-otf**
- Garamond (`GaramondLibre*`): **garamond-libre**
- Garamond Math (`Garamond-Math*`): **garamond-math**
- KP-fonts (`kp*`): **kpfonts-otf**
- Libertine (`LinBiolinum*`, `LinLibertine*`): **libertine**
- TeX Gyre (`texgyre*`): **tex-gyre**
- TeX Gyre Math (`texgyre-math*`): **tex-gyre-math**


# Conclusion

Large margins are optimal with the `companion` style and the `verbose-inote` bibliographic style; small margins should be used with the default options of the template.
Much of memoir customization is offered by the template by choosing other styles, and possibly other fonts.
