---
title: "Tables with memoiR"
output: 
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    number_sections: false
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
# Installation of packages if necessary
install_packages <- function(packages) {
  install_package <- function(package) {
    if (!package %in% installed.packages()[, 1]) {
      install.packages(package, repos = "https://cran.rstudio.com/")
    }
  }
  invisible(sapply(packages, install_package))
}
install_packages(c("gt", "huxtable"))
knitr::opts_chunk$set(cache = FALSE)
set.seed(97310)
```

The purpose here is to print a dataframe as a formated table.

Many packages are dedicated to tables but few of them meet the requirements of memoiR:

- Support of numbered captions for cross-references (such as "Table 1: caption").
- Support of HTML, LaTeX (ultimately PDF) and hopefully Word outputs.
- Support of 2-column LaTeX documents.
- Support of maths in all formats.

**kableExtra** is memoiR's first choice because it supports LaTeX better than the others.
**huxtable** is an alternative if Word support is needed.


# kableExtra

## Basic use

**bookdown** was built with the `kable()` function[^1], completed by the **kableExtra** package.

[^1]: <https://bookdown.org/yihui/rmarkdown-cookbook/kable.html>

This basic example shows how to produce a simple, well-formatted table.
The label of the code chunk is `cars`.
```{r}
library("kableExtra")
mtcars[1:5, 1:6] |>
  kbl(
    caption = "Demo table", 
    booktabs = TRUE
  ) |>
  kable_styling()
```

The `kableExtra::kbl()` function replaces `knitr::kable()` starting from version 1.2 of **kableExtra** to avoid incorrect caption format.
A consequence is that the table caption can't be included in the chunk header with option `tab.cap` which works only with `kable()`.
The table caption is thus in the argument `caption` of `kbl()`.
Formatted captions, possibly with maths, can be used with text references[^5]:

- add the reference somewhere in the text (just above the code chunk is a good place), such as
```
(ref:cars) Caption with _italics_ and maths: $\pi$
```
- call the reference in the chunk with `caption = "(ref:cars)"`.
Note that **knitr** shows the value of the text reference in the chunk below.

(ref:cars) Caption with _italics_ and maths: $\pi$.
```{r}
#| label: cars
library("kableExtra")
mtcars[1:5, 1:6] |>
  kbl(
    caption = "(ref:cars)", 
    booktabs = TRUE
  ) |>
  kable_styling()
```

Argument `booktabs = TRUE` provides clean horizontal rules and should always be used.

Last, `kable_styling()` formats the table for printing.
It has many arguments that are not detailed here: read the documentation of the package[^2].
More details are given in the document templates of memoiR, including the code to place tables in a single column or across the full page in the Stylish Article template.

[^2]: <https://haozhu233.github.io/kableExtra/>
[^5]: <https://bookdown.org/yihui/bookdown/markdown-extensions-by-bookdown.html#text-references>

Cross references (such as table \@ref(tab:cars)) are called by `table \@ref(tab:cars)`: the name of the code snippet ("cars" here) is prefixed by `tab:`.


## Limitations

**kableExtra** used to produce clean tables in Word with a workaround[^3] that was included in previous versions of memoiR but this no longer the case since its version 1.
Tables are rendered with a single cell per row.

[^3]: <https://github.com/haozhu233/kableExtra/issues/308>

The solution is to produce the HTML version of the document, copy its tables and paste them into the Word version[^4].

[^4]: <https://haozhu233.github.io/kableExtra/kableExtra_and_word.html>


# huxtable

## Basic use

**huxtable** works perfectly with memoiR after adding the LaTeX packages in the `preamble` section of the YAML header.
Run `huxtable::report_latex_dependencies()` and copy all of them to obtain something like

```
preamble: >
  \usepackage{array}
  \usepackage{caption}
  ...
```
Add **huxtable** in the vector of packages to install in the "Options" chunk:

```{r}
#| eval: false
# Add necessary packages here
packages <- c("tidyverse", "huxtable")
```

Tables are produced by `as_hux()` and formatting is made possible by a large set of functions.

(ref:ht) Made by **huxtable**
```{r}
#| label: ht
#| message: false
library("huxtable")
mtcars[1:5, 1:6] |>
  as_hux() |>
  set_header_rows(1, value = TRUE) |> 
  style_headers(bold = TRUE)|> 
  theme_article() |>
  set_caption("(ref:ht)")
```

Read its documentation[^6].

[^6]: <https://hughjonesd.github.io/huxtable/reference/>

## Limitations

Word output is correct in contrast with **kableExtra** but LaTeX support is yet more limited.
In Stylish Articles, tables are always printed in full width.
There seems to be no way to make them fit a single column.


# gt

**gt** is the a promising table package, developped by the RStudio team with an approach similar to that of **ggplot2**, i.e. the definition of a grammar of tables.

It has been focusing on HTML output and somehow neglected LaTeX in its first versions.
LaTeX support was improved in version 0.11 but it can't be used in memoiR yet because the numbered captions are not correct, even in HTML, as shown below.

```{r}
#| label: gt
library("gt")
mtcars[1:5, 1:6] |>
  gt(
    caption = "Demo table"
  )
```


# flextable

**flextable** supports all output formats but it does not produce standard table captions.
Even though they are numbered and support cross-referencing, they rely on their own style, which makes them different from figure captions.
This requires extra work to harmonize the documents.

More problematic, knitting to PDF fails with memoiR templates because the LaTeX templates are not compatible.

In conclusion, flextable is not supported.
